\vspace{-2cm} Inference on the Effects of Observed Features \\ in Latent Space Models for Networks

### 2017-04-03 response.tex

\subsection{Response to Reviewer 1}\label{response-to-reviewer-1}

\begin{quote}
``The paper does not accurately characterize or analytically understand
the latent space approaches employed by the work that is cited. For
example Dorff and Ward (2013) do not use LATENTNET. LATENTNET is
extremely outdated and is not the software used in much of the other
work cited herein. This undermines much of the criticisms in the papers
and raises a red flag that suggests the authors do not grasp the
modeling approaches as deeply as they would present.''
\end{quote}

While we appreciate that the \texttt{latentnet} package is not the
implementation most commonly used in political science, it is used
outside of political science. We have made this point more clearly in
the revised manuscript. Most importantly, our criticism is not specific
to the implementation in \texttt{latentnet}, and we argue (again, more
clearly in the revised manuscript), that our argument does not depend on
the specific software implementation or parameterization used, and hence
this criticism does not speak to our point, which is that that the LSM
cannot adjust for unobserved confounders, even in a synthetic scenario
designed to make this as easy as possible.

We updated the Problem section of the manuscript to represent, (1) that we know about extensions to the LSM, (2) to be clear that we think the problem affects all versions of the LSM, and (3) to quote from AMEN to echo the problem we see with using LSM as a method of adjusting for confounders. We cannot possibly analyze all of the individual variants of the LSM, but our point is more broadly directed at the use of latent variables to adjust for confounding network dependencies.

@@

\begin{quote}
``The authors are not up-to-date on the state of latent modeling in the
field and have ignored more recent work (which addresses many of the
concerns raised by the authors) particularly by Peter Hoff and Shahryar
Minhas. Additionally, the progress made by the newly developed AMEN
software is completely overlooked by this article.''
\end{quote}

We have added discussion of the AMEN package, which we were aware of,
but again, does not apply to our argument for the same reason.

\begin{quote}
``In that vein, authors identify a problem (e.g., collinearity between
latent distance parameters and exogenous variables in the latentnet
framework) and do not solve it nor discuss any recent work that tries to
address this problem.''
\end{quote}

We added discussion of the way in which AMEN addresses the problem of correlation between the observed covariates and the latent factors. We point out, however, that the AMEN approach is designed to address Type 2 error by prioritizing the observed covariates and using the latent factors to explain the variation in the residuals. This does not address Type 1 error (which is more troubling than Type 2), and probably makes Type 1 worse.

### 2017-05-03 manuscript.tex
amen added to manuscript

This ``Euclidean'' LSM is the original form proposed by \citet{hoff2002latent}, and, as far as we can tell, the most commonly used specification in the literature. However, the LSM has been extended in several important directions. The  The latent space has been represented as a $k$-dimensional Euclildean space and by latent factors with a multiplicative metric \citep{hoff2002latent, hoff2004modeling, hoff_2005_jasa, hoff2009multiplicative}. Additional structure has been introduced by adding random effects, which, for example, may involve sender or receiver specific effects for directed networks which capture differential activity rates amongst nodes \citep{hoff2003random}. Within this framework \cite{westveld2011mixed} models dynamic network data by treating the latent space as a stochastic process. \cite{handcock2007model} enable the LSM to model clustering that is not representable as homophily (i.e., stochastic equivalence) by combining the LSM with latent cluster models. \cite{hoff2008modeling} shows that the LSM and latent cluster models are special cases of an ``eigenmodel.'' That is, an eigendecomposition of a symmetric sociomatrix can be used to represent both latent space and cluster models, but not vice-versa. \citet{hoff2015dyadic} consolidates, extends, and implements several of these extensions in the \texttt{amen} package \citep{amen} for \R.

@@

Recent work on the LSM has explicitly addressed the potential problem of Type 2 error. The AMEN framework for latent variable modeling of networks is explicitly designed to model structure in the residuals---structure that is leftover after accounting for the observed covariates. \citet[p. 43]{hoff2015dyadic} notes that the latent factors represent patterns in the network that, ``aren't explained by the known regressors.'' \citet[pp. 12--13]{minhas2016inferential} also describe how AMEN is designed such that the multiplicative effects, ``capture higher-order dependence patterns that are left over [in the stochastic linear predictor] after accounting for any known covariate information.''  The approach incorporated in AMEN is effective at avoiding correlation between the observed covariates and latent factors, which would result in Type 2 error. However, there have been no methodological innovations to avoid Type 1 error in latent variable modeling with networks. Below we review several studies in which researchers have relied on the latent space modeling framework to adjust for unmeasured confounders, and through a simulation study we show that the LSM performs poorly in terms of Type 1 error when the unmeasured network structure confounds the relationship between the observed covariate and the network.

@@

@@

@@

Based on our simulation study, we conclude that the primary advantage of using the LSM---relative to a dyadic regression model in which the network is fit using observed covariates only---is that the latent space provides an efficient complement to observed covariates when it comes to fitting and predicting the network. Considering both (1) that the LSM does not substantially reduce bias or inferential error relative to the GLM, and (2) that the LSM exhibits strong predictive performance, we conclude that the latent positions are used to explain the systematic variation that cannot be explained through the observed covariates. This is of considerable use in research focused on developing a predictive model, or fitting and exploring latent positions. However, since the latent positions are inferred to complement the observed covariates---explaining residual variation---the LSM is ill-suited to adjust for confounding network structure, as adjusting for confounding would require that the latent positions explain variation that could otherwise be attributed to the observed covariates.

>>>>

This ``Euclidean'' LSM is the original form proposed by \citet{hoff2002latent}, and, as far as we can tell, the most commonly used specification in the literature. However, the LSM has been extended in several important directions. The  The latent space has been represented as a $k$-dimensional Euclildean space and by latent factors with a multiplicative metric \citep{hoff2002latent, hoff2004modeling, hoff_2005_jasa, hoff2009multiplicative}. Additional structure has been introduced by adding random effects, which, for example, may involve sender or receiver specific effects for directed networks which capture differential activity rates amongst nodes \citep{hoff2003random}. Within this framework \cite{westveld2011mixed} models dynamic network data by treating the latent space as a stochastic process. \cite{handcock2007model} enable the LSM to model clustering that is not representable as homophily (i.e., stochastic equivalence) by combining the LSM with latent cluster models. \cite{hoff2008modeling} shows that the LSM and latent cluster models are special cases of an ``eigenmodel.'' That is, an eigendecomposition of a symmetric sociomatrix can be used to represent both latent space and cluster models, but not vice-versa. \citet{hoff2015dyadic} consolidates, extends, and implements several of these extensions in the \texttt{amen} package \citep{amen} for \R \citep{minhas2016inferential, hoff2015dyadic}.

@@

Recent work on the LSM has explicitly addressed the potential problem of Type 2 error. The AMEN framework for latent variable modeling of networks is explicitly designed to model structure in the residuals---structure that is leftover after accounting for the observed covariates. \citet[p. 43]{hoff2015dyadic} notes that the latent factors represent patterns in the network that, ``aren't explained by the known regressors.'' \citet[pp. 12--13]{minhas2016inferential} also describe how AMEN is designed such that the multiplicative effects, ``capture higher-order dependence patterns that are left over [in the stochastic linear predictor] after accounting for any known covariate information.''  The approach incorporated in AMEN is effective at avoiding correlation between the observed covariates and latent factors, which would result in Type 2 error. However, there have been no methodological innovations to avoid Type 1 error in latent variable modeling with networks. Below we review several studies in which researchers have relied on the latent space modeling framework to adjust for unmeasured confounders, and through a simulation study we show that the LSM as implemented in the R package \texttt{latentnet} as well as \texttt{amen} performs poorly in terms of Type 1 error when the unmeasured network structure confounds the relationship between the observed covariate and the network.

@@

We also repeat this simulation study using \texttt{amen} \citep{hoff2015dyadic, minhas2016inferential}. We omit the poisson family, and, due to the computational cost of the simulation, we consider only networks with 25 nodes or 100 nodes. Rather than constructing an omitted variable as described above, we draw from a standard normal distribution $\mathbf{u}$, and, because we are considering symmetric networks, take the outer product of $\mathbf{u}$ with itself to construct the latent factor. We take 100,000 MCMC samples after x burn-in iterations, and we assess convergence using the diagnostic of \cite{raftery1992practical}.

@@

In our repetition of this simulation with \texttt{amen} we find similar results, with the exception of type-2 error rates, which, as expected, are substantially lower for \texttt{amen}. However, type-1 error rates remain elevated, as is bias in the case where there is a latent dependence structure that is correlated with the observed covariate. \texttt{amen} does perform better in the case where it is used despite the absence of such a latent dependence structure though, relative to \texttt{latentnet}, which is to be expected given that it outperforms \texttt{latentnet} in terms of type-2 error. Again, \texttt{amen} performs well in terms of generalization error, coming close to the theoretical lower bound.

@@

Based on our simulation study, we conclude that the primary advantage of using the LSM---relative to a dyadic regression model in which the network is fit using observed covariates only---is that the latent space provides an efficient complement to observed covariates when it comes to fitting and predicting the network; \texttt{amen} increases this advantage.
Considering both (1) that the LSM does not substantially reduce bias or inferential error relative to the GLM, and (2) that the LSM exhibits strong predictive performance, we conclude that the latent positions are used to explain the systematic variation that cannot be explained through the observed covariates. This is of considerable use in research focused on developing a predictive model, or fitting and exploring latent positions. However, since the latent positions are inferred to complement the observed covariates---explaining residual variation---the LSM is ill-suited to adjust for confounding network structure, as adjusting for confounding would require that the latent positions explain variation that could otherwise be attributed to the observed covariates.
